Indicators for Ecosystem Condition in Norway

### 2023-06-22 DRAFT_ndvi_wetland.Rmd
ndvi data handling and regression fixes

# need to split the ninbeskrivelsesvariabler-column into one per variable and these then into two for the variable name and value
levels(as.factor(nin$ninbeskrivelsesvariabler))

nin2 <- nin %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
  separate(col = ninbeskrivelsesvariabler,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  mutate(NiN_variable_code = as.factor(NiN_variable_code))

nin2$NiN_variable_code <- as.factor(paste0("var_", nin2$NiN_variable_code, "_end",sep=""))

nin3 <- nin2 %>%
  pivot_wider(
    names_from = "NiN_variable_code",
    values_from = "NiN_variable_value"
  )
# this does not do what it is expected to do. nrow(nin3) should go down towards nrow(nin), but it stays at almost nrow(nin2)
# continue with nin for now, ignoring the beskrivelsesvariabler and only using information from 'tilstand'
rm(nin2)
rm(nin3)

@@

@@

### Regression analyses, NDVI as a function of condition, Sentinel-2
```{r}
# NDVI data are bound between -1 and 1, and thus require modelling with an appropriate method that handles bounded data
# we can transform the variable to be bounded between 0 and 1 and use beta-regression models
SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
  filter(year == kartleggingsaar)

summary( betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train) )
# full model does not work, remove the 3-way interaction
summary( betareg(mean_beta~tilstand_num + region + hovedtype + tilstand_num:region + tilstand_num:hovedtype + region:hovedtype, data=SentinelNDVI.wetland.train) )
# this works! Let's run a stepwise-function to fin the most parsimonious model
model.wetland.train <- betareg(mean_beta~tilstand_num + region + hovedtype + tilstand_num:region + tilstand_num:hovedtype + region:hovedtype, data=SentinelNDVI.wetland.train)
model.wetland.train <- StepBeta(model.wetland.train)
summary(model.wetland.train)
```

@@

Since the 3-way interaction term was not possible to be evaluated by the model, but it is evident that all three variables matter, we run the model separately for every region
```{r}
model.wetland.train.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
model.wetland.train.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
model.wetland.train.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
model.wetland.train.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
model.wetland.train.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])

>>>>

##############################
##############################
# code on ninbeskrivelsesvariabler here if relevant
##############################
##############################

@@

## looking at ndvi in different condition classes, separately for main types and regions  (only for NDVI data matching NiN-mapping)
SentinelNDVI.wetland %>%
  filter(year != '2022') %>%

  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE)) %>%

  filter(year == kartleggingsaar) %>%
  ggplot( aes(x=tilstand, y=mean )) + 
  geom_violin() +
  facet_grid( region~hovedtype)
# V4 has lacking data for quite some condition classes in most regions

@@

### Regression analyses, NDVI as a function of condition, Sentinel-2
```{r}
# NDVI data are bound between -1 and 1, and thus require modelling with an appropriate method that handles bounded data
# we can transform the variable to be bounded between 0 and 1 and use beta-regression models
SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
# we drop V4 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues

SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
  filter(year == kartleggingsaar) %>%
  filter(hovedtype != 'V4')

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
model.wetland.train <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
model.wetland.train <- StepBeta(model.wetland.train)
summary(model.wetland.train)
```

@@

It is evident that all three variables matter, for ease of interpretation we run the model separately for every region
```{r}
model.wetland.train.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
model.wetland.train.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
model.wetland.train.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
model.wetland.train.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
model.wetland.train.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])
